Probabilistic Model of Bilateral Lymphatic Spread in Head and Neck Cancer

### 2024-06-17 manuscript.qmd
add some refs & write first abstract draft

---
title: "Probabilistic Model of Bilateral Lymphatic Spread in Head and Neck Cancer"
---

@@

We have collected a detailed dataset of 833<!-- num patients --> patients with newly diagnosed oropharyngeal squamous cell carcinomas. It reports the involvement of every patient individually and per lymph node level in tabular form, in addition to other clinico-pathological information such as age, T-category, and HPV p16 status.

@@

It is possible to provide our model with multiple different diagnostic modalities, each being characterized by different pairs of sensitivity and specificity. However, we instead chose to combine them into a single "consensus" diagnosis before parameter inference. We opted for this because the literature values of sensitivity and specificity [@debondt_detection_2007] of imaging modalities like MRI and CT do not plausibly match some of our observations: In the USZ cohort, 78% of OPSCC patients where diagnosed with ipsilateral LNL II involvement via diagnostic imaging. This is virtually impossible with sensitivities around 80% and specificities lower than 100%.

Our pre-training consensus was formed by considering all reported diagnostic information for a particular patient and LNL. When conflicts arose, we computed the *most likely* true state of involvement using the literature sensitivity and specificity values [@debondt_detection_2007].

@@

For parameter inference, we used the Python package [`emcee`] [@foreman-mackey_emcee_2013]. It implements efficient MCMC sampling algorithms that employ multiple parallel samplers for affine invariance and better performance on multi-core CPUs. The [`emcee`] library was provided with the likelihood implemented by our [`lymph-model`] Python package.

>>>>

---
title: "Probabilistic Model of Bilateral Lymphatic Spread in Head and Neck Cancer"
abstract: |
  In current practice, large parts of the contralateral lymph system are included in the elective clinical target volume (CTV-N) of oropharyngeal squamous cell carcinoma (OPSCC) patients. Guidelines on how to define the CTV-N are mostly based on prevalences of nodal involvement in the different lymph node levels (LNLs) and do not consider the detailed, individual diagnosis of a patient.

  In this work, we present a probabilistic model for bilateral lymphatic tumor progression in OPSCC. Its aim is to predict the personal risk for occult disease in any LNL, given the patient's clincal diagnosis, T-category, and lateralization of the primary tumor. The model represents each of the LNLs I, II, III, IV, V, and VII of both sides of the neck as a hidden binary random variable that are connected to the tumor and among each other via arcs that correspond to spread probabilities per time-step. These spread probability rates are learned via Markov chain Monte Carlo (MCMC) sampling from a dataset of 833 OPSCC patients.

  The model is able to describe the data well and may in the future be used to only include LNLs in a CTV-N that show a risk above a specific threshold for that particular patient's diagnosis. E.g., it predicts the risk for occult disease in contralateral LNL II to be below 5% unless the tumor crosses the mid-sagittal line.
---

@@

Originally, the only available data of detailed - i.e. per patient and per LNL - lymphatic involvement was a reconstructed table from the findings in @sanguineti_defining_2009, which included only ipsilateral nodal involvement for the LNLs I to IV.

To be able to create models for lymphatic tumor progression that include more LNLs and the contralateral side, we have collected a detailed dataset of 833<!-- num patients --> patients with newly diagnosed oropharyngeal squamous cell carcinomas. It reports the involvement of every patient individually and per lymph node level in tabular form, in addition to other clinico-pathological information such as age, T-category, and HPV p16 status.

@@

It is possible to provide our model with multiple different diagnostic modalities, each being characterized by different pairs of sensitivity and specificity. However, we instead chose to combine them into a single "consensus" diagnosis before parameter inference. We opted for this because the literature values of sensitivity and specificity [@debondt_detection_2007;@kyzas_18ffluorodeoxyglucose_2008] of imaging modalities like MRI and CT do not plausibly match some of our observations: In the USZ cohort, 78% of OPSCC patients where diagnosed with ipsilateral LNL II involvement via diagnostic imaging. This is virtually impossible with sensitivities around 80% and specificities lower than 100%.

Our pre-training consensus was formed by considering all reported diagnostic information for a particular patient and LNL. When conflicts arose, we computed the *most likely* true state of involvement using the literature sensitivity and specificity values [@debondt_detection_2007;@kyzas_18ffluorodeoxyglucose_2008].

@@

For parameter inference, we used the Python package [`emcee`] [@foreman-mackey_emcee_2013]. It implements efficient MCMC sampling algorithms that employ multiple parallel samplers for affine invariance and better performance on multi-core CPUs. The sample proposal algirothms used by us are based on differential evolution moves [@terbraak_differential_2008;@nelson_run_2013]. The [`emcee`] library was provided with the likelihood implemented by our [`lymph-model`] Python package.
